---
title: "Allocating Treatments in a CRD"
#author: "Cécile Ané"
#date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA, fig.height=4, fig.width=9)
# options(show.signif.stars=FALSE)
knitr::opts_knit$set(global.par = TRUE, html.table.attributes = '')
```

```{r, include=FALSE}
par(mar=c(2.1,2.1,.5,.5), mgp=c(1.3,.3,0), tck=-0.01)
set.seed(1)
```

Imagine 4 treatments and 16 units total.
In a completely randomized design (CRD), we want to allocate
4 units to each of the 4 treatments at random:

```{r}
trts = sample(rep(1:4, each=4)) # "rep"eats 1,2,3,4, 4 times each: 1,1,1,1,2,2,...,4
trts # "sample": like drawing from a hat, without replacement
```

now we can arrange these allocations of treatments to units 
in a 4x4 grid:

```{r}
crd = matrix(trts, 4, 4)
```
```{r, echo=FALSE}
crd
```

Here is a "fair" arrangement of the treatments: each
treatment appears only once in each row and column:

<!-- fair = matrix(c(3,2,1,4, 4,1,3,2, 2,3,4,1, 1,4,2,3),
              4,4) # in 4 rows, 4 columns-->
```{r, message=FALSE, results="hide"}
library(magic) # the function "rlatin" is in package "magic"
fair = rlatin(4)
```

```{r, echo=FALSE}
fair
```

Now suppose we have some data, with a hidden trend.
In a corn field, this trend might be caused by heterogeneity,
in the *absence* of any treatment:
the plots on the left might be well watered,
and the plots on the right might be too dry (uphill):

```{r}
dat = c(12,14,15,17,  7,6,10,12,  5,4,8,7,  4,3,2,1)
matrix(dat, 4,4) # to view in 4 rows, 4 columns
```

Suppose there is no treatment effect.  Then we are really
just randomly assigning the labels "1", "2", "3", and "4" to
the plots. From each (random) assignment, we can calculate
the F test statistic. It should look like what it does when
the null hypothesis is true: an F distribution.

```{r, include=FALSE}
set.seed(2)
```

```{r simulrandom, height=4, width=9}
n_simulations = 1000
FstatRandom = rep(NA, n_simulations) # 1000 missing values for now
FstatFair = rep(NA, n_simulations)
for (i in 1:n_simulations){
  trts = sample(rep(1:4, each=4))
  fit  = lm(dat ~ as.factor(trts))
  FstatRandom[i] = anova(fit)$F[1] # i-th F stat: extracted from ANOVA table
  trts = c(rlatin(4)) # do it over, but fair allocation
  fit  = lm(dat ~ as.factor(trts))
  FstatFair[i] = anova(fit)$F[1]
}
layout(matrix(1:2,1,2))
hist(FstatRandom, probability=TRUE, breaks=40, ylim=c(0,.8), xlim=c(0,10),
     col="tan", main="", xlab="F statistic")
xx = seq(0, 10, 0.01)
lines(xx, df(xx, 3, 12)) # df = density of F distribution
hist(FstatFair, probability=TRUE, breaks=2, xlim=c(0, 4), col="tan", main="")
lines(xx, df(xx, 3, 12))
```

above left: histogram of `r n_simulations` F-statistics based on
`r n_simulations` random allocations of treatment labels in a CRD
with k=4 treatments and n=4 units per treatments.
The smooth line corresponds to an F<sub>3,12</sub> distribution.

above right: the same but with "fair" allocations, using latin squares.
histogram of `r n_simulations` F-statistics based on
`r n_simulations` "fair" allocations of treatment labels in a
CRD with k=4 treatments and n=4 units per treatments.
The smooth line corresponds to an F<sub>3,12</sub> distribution.

### &nbsp;
<!-- to make space only-->

### latin square analysis: to follow the design

if we assign treatments with a latin square ("fair" assignment),
we should analyze the resulting data accordingly: with a row factor,
a column factor, and the treatment factor, like in this example
(note that the column effect is detected as significant).

```{r}
plotrow = factor(rep(1:4, 4))      # the row number: 1,2,3,4, 1,2,3,4, etc.
plotcol = factor(rep(1:4, each=4)) #  column number: 1,1,1,1, 2,2,2,2, etc.
trts = c(rlatin(4)) # one random fair allocation of treatments
fit  = lm(dat ~ plotrow + plotcol + as.factor(trts))
drop1(fit, test="F")
```

<!--
The simulation with "fair" allocations (above right) was repeated,
but each data set was analyzed like this:

the F-statistics are no longer biased when the analysis follows the design
(full simulation code not shown):
(actually, F takes only 1 of 5 values)

```{r latinsquare, height=4, width=9, echo=FALSE}
n_simulations = 1000
FstatFair = rep(NA, n_simulations) # 1000 missing values for now
for (i in 1:n_simulations){
  trts = c(rlatin(4)) # fair allocation of treatments
  fit  = lm(dat ~ plotrow + plotcol + as.factor(trts)) # latin square model
  FstatFair[i] = anova(fit)$F[1] # extract F value
}
layout(matrix(1:2,1,2))
hist(FstatFair, probability=TRUE, breaks=40, ylim=c(0,.8), xlim=c(0,10), col="tan", main="")
lines(xx, df(xx, 3, 12)) # df = density of F distribution
```
-->